Strict Pure Functional Programming and Automatic Parallelization\\ (New Project-in-progess Paper)

### 2016-02-26 pubs/tfp16/paper.tex
sync before branching

\subsubsection{Rely on programmer discipline.}
Legacy Fortran scientific codes likely represent worst practices in state
encapsulation, with most data global and accessible to all.  It is almost
ironic, then, that of the mainstream high-performance scientific programming
languages (Fortran, C, and C++), Fortran 95 was the first to introduce the
\texttt{pure} function qualifier, and with enforcement by the compiler.
Otherwise it is entirely up to the programmer to make functions composable
and thread safe.  Here at LANL this approach is being used to good effect
to clean up some legacy code bases.

@@

In Legion, non-function-local data, or \emph{regions}, are requested from the
runtime system.  Regions may be shared among \emph{tasks} (essentially C++
functions), and each task registers its access (read, write, read/write, etc.)
and coherency requirements for each region to which it will have access.
Other than registered access to regions tasks must not access non-constant
global data.  In serial execution these access requirements are notionally
superfluous, but in parallel execution the runtime can dynamically calculate
the dependency graph and relax the programmatically defined strict ordering of
task execution.  Very high parallel efficiency has been demonstrated on extremely
large computing systems~\cite{Bauer14}.  Here at LANL the Legion system is
being strongly considered for a new large-scale C++ code effort.\footnote{
It is interesting to note that the PI of the Legion project was previously a
co-developer of a strict, arguably pure functional language
implementation~\cite{AikenFL}.}

\subsubsection{Use a pure functional language.}
Some twenty years after the US Department of Energy laboratory complex was
presented with the most performant and highly parallel implementation of a
pure functional language to date~\cite{Davis96}, and collectively had no
conceptual understanding of what it was, much less why it might be useful or
important, there is a small number of laboratory HPC practioners who are
developing an appreciation for pure functional programming in Haskell.  Still,
they see their needs as different than other Haskell users: performance is
paramount, unboxed arrays are a (perhaps \emph{the}) primary data structure,
and laziness is more of a nuisance--even a serious hindrance--than a
convenience, much less algorithmically essential in the sense of Bird's
\emph{repmin}~\cite{Bird84} or Johnsson's general attribute
grammars~\cite{Johnsson87}.

A common complaint among these budding high-performance functional programmers
is the need to get strictness just right: to remove space leaks,
improve performance, get communication (e.g.\ MPI) and IO to work as they
expect, and so on.  More specifically, they expect to be able to reason about
space and time usage using their existing mental models. While they are
happy to write code that is strongly reminiscent of the numerical examples in
Hughes' classic piece (whether they've read it or not)~\cite{Hughes89},
they would also be happy to write in a slightly modified style that does
not require non-strictness---after all, that's how they've been thinking their
entire careers.

@@

\begin{itemize}
\item The implementation must be feasible in terms of effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{firstclass};
\item Straighforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
\item The language should be ``currying-friendly'' as described in the ``fast
  curry'' paper~\cite{fastcurry};
\item Choice of degree of strictness is easily selected;
\item Sharing, in the sense of what ``lazy'' means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item Tail calling should be properly implemented, i.e., not as a trampoline;
\item etc.
\end{itemize}

@@

First some words about terminology.  For all degrees of strictness sharing (as
implied by lazy but not by non-strict) of heap-allocated objects is preserved
we will not use the term \emph{lazy} and simply refer to strictness
properties.  As is common we will use strict in an operational sense, meaning
that argument(s) are evaluated before being passed in a function call.
Similarly, when referring to language semantics we refer to the operational
rather than denotational semantics.

@@

ghc 8.0 gives three possibilities: conventional non-strict, constructor
strict, and function- and constructor strict, but there are possible
variations.  The obvious variation is function, but not constructor, strict.
In Haskell the distinction between constructors and functions is somewhat
blurred because constructors behave like functions (outside of pattern
matching), but this need not be the case.  In STG---ghc's intermediate
language---the distinction is clear: heap objects are explicitly constructed
by \texttt{let} constructs in which constructor names are explicitly mentioned
(not aliased); constructor names are not functions.

Other possible variations concern the notional arity of the underlying
function being applied.  This is readily explained in terms of the operation of the
STG machine described next.

This then gives three reasonable operational modes that are strictly ordered
in terms of termination properties: non-strict, fully strict---all arguments
in an application are evaluated---and the intermediate in which only those
arguments needed by a function call are evaluated.

@@

Variable     & $f,\ x$        &     &                                              & Initial lower-case letter \\
Constructor  & $C$            &     &                                              & Initial upper-case letter \\
Literal      & $\mathit{lit}$ & ::= & $i\ |\ d$                                    & Integral or floating point \\
Atom         & $a$            & ::= & $\mathit{lit}\ |\ x$                         & \\ % Function, constructor args are atoms \\
\\
Expression   & $e$            & ::= & $a$                                          & Atom \\
             &                & $|$ & $f\ a_1\dots a_n$                            & Function application, $n\ge 1$ \\
             &                & $|$ & $\oplus\ a_1\dots a_n$                       & Saturated primitive operation, $n\ge 1$ \\
             &                & $|$ & \texttt{let} $x_i = \mathit{obj}_i$ 
                                         \texttt{in} $e$                           & Recursive let $1\le i \le n$\\
             &                & $|$ & \texttt{case} $e$ \texttt{as} 
                                        $x$ \texttt{of} $\mathit{alt}_i$           & Case expression, $1\le i \le n$\\
\\
Alternatives & $\mathit{alt}$ & ::= & $C\ x_1\dots x_n$ \texttt{->} $e$            & Pattern match, $n \ge 0$ \\
             &                & $|$ & \texttt{\_ ->} $e$                           & Default \\
\\
Heap objects & $\mathit{obj}$ & ::= &\texttt{FUN} $x_1\dots x_n$ \texttt{->} $e$   & Function definition, arity $=n\ge 1$ \\
             &                & $|$ &\texttt{CON} $C\ a_1\dots a_n$                & Saturated constructor, $n \ge 0$ \\
             &                & $|$ &\texttt{THUNK} $e$                            & Thunk \\
             &                & $|$ & $\mathit{PAP}\ f\ a_1\dots a_n$               & Evaluation-time partial application \\
             &                & $|$ & $\mathit{BLACKHOLE}$                         & Evaluation-time black hole \\
\\
Program      & $\mathit{prog}$& ::= & $f_i\ =\ \mathit{obj}_i$                      & $n\ge 1$, distinguished $\mathit{main}$ \\

@@

While our interpretation of STG is a higher-order, polymorphic, pure functional
language, some syntactic restrictions make direct programming rather tedious.
This is discussed further in a later section.

@@

\subsection{Strictness}

The STG machinery makes the choice of evaluation strategy easy to change, both
in terms of function and constructor application.

>>>>

\subsubsection{Rely on programmer discipline.}
Legacy Fortran scientific codes likely represent worst practices in state
encapsulation, with most data global and accessible to all; indeed, Fortran
was designed to make this easy via \texttt{COMMON} blocks---aggregates of
global variables---which further obfuscate program meaning by allowing
arbitrary naming of a block's variables at a subroutine/function granularity,
and requiring that the entire block be brought into scope.  It is almost
ironic, then, that of the mainstream high-performance scientific programming
languages (Fortran, C, and C++), Fortran~95 was the first to introduce the
\texttt{pure} function qualifier, and with enforcement by the compiler.
Otherwise it is entirely up to the programmer to make functions composable and
thread safe.  Here at LANL this approach is being used to good effect to clean
up some legacy code bases.

@@

In Legion, non-function-local data, or \emph{regions}, are requested from the
runtime system.  Regions may be shared among \emph{tasks} (essentially C++
functions), and each task is prescribed access privileges (read, write,
read/write, etc.)  and coherency requirements for each region to which it will
have access.  Other than registered access to regions tasks must not access
non-constant global data.  In serial execution these access requirements are
notionally superfluous, but in parallel execution the runtime can dynamically
calculate the data dependency graph and relax the programmatically defined
serial ordering of task execution.  Very high parallel efficiency has been
demonstrated on extremely large computing systems~\cite{Bauer14}.  Here at
LANL the Legion system is being strongly considered for a new large-scale C++
code effort.\footnote{It seems significant that the PI of the Legion
  project was previously a co-developer of a strict, arguably pure functional
  language implementation~\cite{AikenFL}.}

\subsubsection{Use a pure functional language.}
Some twenty years after the US Department of Energy laboratory complex was
presented with the most highly parallel (and thereby performant)
implementation of a pure functional language to date~\cite{Davis96}, and
collectively had no conceptual understanding of what it was, much less why it
might be useful or important, there is now a small number of laboratory HPC
practioners who are developing an appreciation for pure functional programming
in Haskell.  Still, they see their needs as different than other Haskell
users: performance is paramount, unboxed arrays are a (perhaps \emph{the})
primary data structure, and laziness is more of a nuisance--even a serious
hindrance--than a convenience, much less ever algorithmically essential in the
sense of Bird's \emph{repmin}~\cite{Bird84} or Johnsson's general attribute
grammars~\cite{Johnsson87}.

A common complaint among these budding high-performance functional programmers
is the need to get strictness just right: to remove space leaks, achieve
acceptable performance, get communication (e.g.\ MPI) and IO to work as they
expect, and so on.  More specifically, they expect to be able to reason about
space and time usage using their existing mental models. While they are happy
to write code that is strongly reminiscent of the numerical examples in
Hughes' classic piece (whether they've read it or not)~\cite{Hughes89}, they
would also be happy to write in a slightly modified style that does not
require non-strictness---after all, that's how they've been thinking their
entire careers.

@@

\begin{itemize}
\item The implementation must be feasible in terms of effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{firstclass};
\item Straighforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
\item The language should be \emph{currying-friendly} as described in the ``fast
  curry'' paper~\cite{fastcurry};
\item Choice of degree of strictness is easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item Tail calling should be properly implemented, i.e., not as a trampoline;
\item etc.
\end{itemize}

@@

First some words about terminology.  For all degrees of strictness, sharing
(as implied by lazy but not by non-strict) of heap-allocated objects is
preserved so we will not use the term \emph{lazy} and simply refer to
strictness properties.  As is common we will use strict in an operational
sense, meaning that argument(s) are evaluated before being passed in a
function call, rather than the denotational sense.  Similarly, when referring
to language semantics we refer to the operational rather than denotational
semantics.

@@

ghc 8.0 gives three possibilities: conventional non-strict, constructor
strict, and function- and constructor strict, but there are possible
variations.  The obvious variation is function, but not constructor, strictness.
In Haskell the distinction between constructors and functions is somewhat
blurred because constructors behave like functions (outside of pattern
matching), but this need not be the case.  In STG---one of ghc's intermediate
representations---the distinction is clear: heap objects are explicitly constructed
by \texttt{let} constructs in which constructor names are given explicitly
(not aliased); constructor names are not functions.

@@

Variable     & $f,\ x$        &     &                                              & Initial lower-case letter \\
Constructor  & $C$            &     &                                              & Initial upper-case letter \\
Literal      & $\mathit{lit}$ & ::= & $i\ |\ d$                                    & Integral or floating point literal \\
Atom         & $a$            & ::= & $\mathit{lit}\ |\ x$                         & \\ % Function, constructor args are atoms \\
\\
Expression   & $e$            & ::= & $a$                                          & Atom \\
             &                & $|$ & $f\ a_1\dots a_n$                            & Application, $n\ge 1$ \\
             &                & $|$ & $\oplus\ a_1\dots a_n$                       & Saturated primitive operation, $n\ge 1$ \\
             &                & $|$ & \texttt{let} $x_i = \mathit{obj}_i$ 
                                         \texttt{in} $e$                           & Recursive let $1\le i \le n$\\
             &                & $|$ & \texttt{case} $e$ \texttt{as} 
                                        $x$ \texttt{of} $\mathit{alt}_i$           & Case expression, $1\le i \le n \ge 1$\\
\\
Alternatives & $\mathit{alt}$ & ::= & $C\ x_1\dots x_n$ \texttt{->} $e$            & Pattern match, $n \ge 0$ \\
             &                & $|$ & \texttt{\_ ->} $e$                           & Default \\
\\
Heap objects & $\mathit{obj}$ & ::= &\texttt{FUN} $x_1\dots x_n$ \texttt{->} $e$   & Function definition, arity $=n\ge 1$ \\
             &                & $|$ &\texttt{CON} $C\ a_1\dots a_n$                & Saturated constructor, $n \ge 0$ \\
             &                & $|$ &\texttt{THUNK} $e$                            & Thunk---explicit deferred evaluation \\
             &                & $|$ & $\mathit{PAP}\ f\ a_1\dots a_n$               & Evaluation-time partial application \\
             &                & $|$ & $\mathit{BLACKHOLE}$                         & Evaluation-time black hole \\
\\
Program      & $\mathit{prog}$& ::= & $f_i\ =\ \mathit{obj}_i$                      & $1\le i \le n \ge 1$, distinguished $\mathit{main}$ \\

@@

While our interpretation of STG is as a simple higher-order, polymorphic, pure
functional language, some syntactic restrictions make direct programming
rather tedious.  This is discussed further in a later section.

\subsection{Degrees of strictness, take two}

Other possible variations concern the notional arity of the underlying
function being applied.  This is readily explained in terms of the operation
of the STG machine described later.  Our description is based on that given in
Marlow and Peyton Jones' \emph{fast curry} paper~\cite{fastcurry}, to which we
refer so many times that we subsequenty refer to it as
\emph{EA}.\footnote{\emph{Eval/Apply} from the title because \emph{FC} likely has another
  connotation to the reader.}

This then gives three reasonable operational modes that are strictly ordered
in terms of termination properties: non-strict, fully strict---all arguments
in an application are evaluated---and the intermediate in which only those
arguments needed by a function call are evaluated.  The STG machinery makes
the choice of evaluation strategy easy to change, both in terms of function
and constructor application.

@@
